---
title: "Getting Started with nonBgfa"
description: "A comprehensive guide to detecting non-B DNA motifs"
author: "Copilot"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    highlight-style: github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

Non-B DNA structures are alternative DNA conformations that deviate from the standard Watson-Crick B-form double helix. These structures—including G-quadruplexes, inverted repeats (cruciforms), direct repeats (slipped-strand), mirror repeats (triplex), A-phased repeats, short tandem repeats, and Z-DNA—have been implicated in:

- Genomic instability and rearrangements
- Mutagenesis and disease mechanisms
- Regulation of gene expression
- DNA replication and recombination

The **nonBgfa** package provides an easy-to-use R interface to detect and characterize these motifs in DNA sequences.

## Installation

### From GitHub

```r
# Install from GitHub (requires devtools)
if (!require("devtools")) {
  install.packages("devtools")
}

devtools::install_github("example/nonBgfa")
```

### System Requirements

- R ≥ 3.5
- C99 compiler (gcc, clang, or MSVC)

## Loading the Package

```{r load, eval=FALSE}
library(nonBgfa)
```

## Basic Usage

### Analyzing a DNA Sequence

The simplest way to analyze a DNA sequence is to pass it directly to `gfa_analyze()`:

```r
# Load the package
library(nonBgfa)

# Analyze a simple DNA sequence
dna <- "ACGTACGTACGTACGTACGTACGTACGTACGTACGTACGTACGTACGTACGT"
results <- gfa_analyze(dna)
```

The function returns a list containing data frames for each motif type detected:

```r
# View summary
print(results)

# Check what motif types were detected
names(results)
```

### Analyzing a FASTA File

For larger sequences or batch analysis, use FASTA files:

```r
# Create a sample FASTA file
fasta_content <- ">sequence_1
ACGTACGTACGTACGTACGTACGTACGTACGTACGTACGTACGTACGTACGT
AAAAAACCCCCGGGGGTTTTTTACGTACGTACGTACGTACGTACGTACGT"

writeLines(fasta_content, "example.fasta")

# Analyze the FASTA file
results <- gfa_analyze("example.fasta")
```

## Understanding the Results

### Result Structure

Each result from `gfa_analyze()` is a list containing data frames for GFF and TSV formats for each motif type:

```r
results <- gfa_analyze("sequence.fasta")

# Available components
str(results, max.level = 1)
```

### Accessing Specific Motif Results

```r
# Access G-Quadruplex results in GFF format
gq_gff <- results$gq_gff
head(gq_gff)

# Access Inverted Repeat results in TSV format
ir_tsv <- results$ir_tsv
head(ir_tsv)

# All motif types
# $apr_gff, $apr_tsv  - A-Phased Repeats
# $dr_gff, $dr_tsv    - Direct Repeats
# $ir_gff, $ir_tsv    - Inverted Repeats (cruciforms)
# $gq_gff, $gq_tsv    - G-Quadruplexes
# $mr_gff, $mr_tsv    - Mirror Repeats (triplex)
# $str_gff, $str_tsv  - Short Tandem Repeats
# $z_gff, $z_tsv      - Z-DNA
```

### Understanding Output Columns

**GFF Format Columns:**
- `seqname`: Sequence/chromosome identifier
- `source`: Data source (always "GFA")
- `feature`: Motif type
- `start`: Start position (1-based)
- `end`: End position (inclusive)
- `score`: Detection score
- `strand`: DNA strand (+ or -)
- `frame`: Reading frame (. for not applicable)
- `attributes`: Additional metadata (key=value pairs)

**TSV Format Columns:**
Extends GFF format with additional annotations:
- All GFF columns, plus
- `islands`: Number of motif islands detected
- `runs`: Number of runs within islands
- `max`: Maximum run length
- `composition`: Nucleotide composition breakdown
- `sequence`: The detected motif sequence

## Customizing Parameters

The default parameters work well for most analyses, but you can customize them using `gfa_params()`:

```r
# Create custom parameters
params <- gfa_params(
  # G-Quadruplex settings
  minGQrep = 4,              # Stricter G-run minimum
  maxGQspacer = 5,           # Tighter spacer constraint
  
  # Inverted Repeat settings
  minIRrep = 8,              # Longer minimum IR
  maxIRspacer = 50,          # Tighter spacer
  
  # Short Tandem Repeat settings
  minSTRbp = 12,             # Longer minimum length
  
  # Skip certain motif searches
  skipAPR = FALSE,           # Search for APR
  skipDR = FALSE,            # Search for DR
  skipGQ = FALSE             # Search for GQ
)

# Analyze with custom parameters
results <- gfa_analyze("sequence.fasta", params = params)
```

### Key Parameters

| Parameter | Default | Range | Description |
|-----------|---------|-------|-------------|
| `minGQrep` | 3 | ≥1 | Minimum G's per G-run |
| `maxGQspacer` | 7 | ≥1 | Max distance between G-runs |
| `minIRrep` | 6 | ≥1 | Min half-length of IR |
| `maxIRspacer` | 100 | ≥0 | Max spacer between IR halves |
| `minDRrep` | 10 | ≥1 | Min half-length of DR |
| `maxDRspacer` | 100 | ≥0 | Max spacer between DR halves |
| `minMRrep` | 10 | ≥1 | Min half-length of MR |
| `maxMRspacer` | 100 | ≥0 | Max spacer between MR halves |
| `minZlen` | 10 | ≥1 | Min Z-DNA length |
| `minSTR` | 1 | ≥1 | Min STR repeat unit length |
| `maxSTR` | 9 | ≥1 | Max STR repeat unit length |
| `minATracts` | 3 | ≥1 | Min A-tract count for APR |

See `?gfa_params` for all 25 parameters.

## Selective Motif Searching

Use skip flags to disable detection of specific motif types:

```r
# Only search for G-Quadruplexes and Z-DNA
results <- gfa_analyze(
  "sequence.fasta",
  skip_apr = TRUE,
  skip_dr = TRUE,
  skip_ir = TRUE,
  skip_mr = TRUE,
  skip_str = TRUE
)

# Only includes results$gq_gff, results$gq_tsv, results$z_gff, results$z_tsv
```

This can improve performance when you're only interested in specific motif types.

## Working with Results

### Filtering Results

```r
# Get G-Quadruplexes with score > 0.5
gq_results <- results$gq_gff
gq_filtered <- gq_results[gq_results$score > 0.5, ]

# Get long inverted repeats (span > 100 bp)
ir_results <- results$ir_gff
ir_long <- ir_results[ir_results$end - ir_results$start > 100, ]
```

### Combining Results Across Motif Types

```r
# Get all detected motifs in one data frame
all_motifs <- do.call(rbind, 
  list(
    results$apr_gff,
    results$dr_gff,
    results$ir_gff,
    results$gq_gff,
    results$mr_gff,
    results$str_gff,
    results$z_gff
  )
)

# Sort by genomic position
all_motifs <- all_motifs[order(all_motifs$start), ]
head(all_motifs)
```

### Exporting Results

```r
# Export to CSV
write.csv(results$gq_gff, "gquadruplexes.csv", row.names = FALSE)
write.csv(results$ir_tsv, "inverted_repeats.csv", row.names = FALSE)

# Export to GFF format (compatible with genome browsers)
write.table(results$gq_gff, "gquadruplexes.gff", 
            sep = "\t", quote = FALSE, row.names = FALSE)
```

## Advanced Examples

### Example 1: Batch Analysis of Multiple Sequences

```r
# Create multiple FASTA files
fasta_files <- c("seq1.fasta", "seq2.fasta", "seq3.fasta")

# Analyze all files
results_list <- lapply(fasta_files, gfa_analyze)

# Combine G-Quadruplex results
all_gq <- do.call(rbind, 
  lapply(results_list, function(x) x$gq_gff)
)

# Count motifs per sequence
table(all_gq$seqname)
```

### Example 2: Compare Parameter Sets

```r
# Default parameters
results_default <- gfa_analyze("sequence.fasta")

# Stricter parameters
params_strict <- gfa_params(
  minGQrep = 5,
  minIRrep = 10,
  minDRrep = 15
)
results_strict <- gfa_analyze("sequence.fasta", params = params_strict)

# Compare results
cat("Default - GQ found:", nrow(results_default$gq_gff), "\n")
cat("Strict - GQ found:", nrow(results_strict$gq_gff), "\n")
```

### Example 3: Focus on Specific Regions

```r
# If sequence is >100 kbp, analyze regions
large_seq <- "ACGT..." # Very long sequence

# Analyze subsequences
results_region1 <- gfa_analyze(substring(large_seq, 1, 50000))
results_region2 <- gfa_analyze(substring(large_seq, 50001, 100000))
```

## Troubleshooting

### Sequence Validation Errors

```r
# Error: Invalid nucleotides
# Solution: Ensure sequence contains only ACGT (and optionally NRYSWKMBDHV)
dna <- "ACGTACGTN"  # OK - N is ambiguous
dna <- "ACGTACGTX"  # ERROR - X is not valid

# Check your sequence
is_valid <- grepl("^[ACGTNRYSWKMBDHV]+$", your_sequence, ignore.case = TRUE)
```

### FASTA File Issues

```r
# Error: File not found
# Solution: Check file path
file.exists("path/to/sequence.fasta")

# Error: Invalid FASTA format
# Solution: Check FASTA formatting
# - First line must start with '>'
# - Sequences follow
# - Can be multiline
```

### Empty Results

```r
# If no motifs detected, this is normal for some parameters
if (nrow(results$gq_gff) == 0) {
  cat("No G-Quadruplexes detected\n")
}

# Try relaxing parameters
params_loose <- gfa_params(minGQrep = 2, maxGQspacer = 10)
results_loose <- gfa_analyze("sequence.fasta", params = params_loose)
```

## Background Information

### What Are Non-B DNA Structures?

DNA is a complex molecule with potential for multiple conformations:

**B-DNA** (Watson-Crick): The standard double helix form (~95% of cellular DNA)

**G-Quadruplexes (GQ)**: Four-stranded structures formed by G-rich sequences
- Found near telomeres and gene promoters
- Implicated in aging and cancer

**Inverted Repeats (IR) → Cruciforms**: Palindromic sequences that form hairpin structures
- Can cause chromosomal rearrangements
- Associated with fragile sites

**Direct Repeats (DR) → Slipped-Strand**: Repetitive sequences prone to strand slippage
- Cause microsatellite expansion/contraction
- Lead to Huntington's disease, fragile X syndrome

**Mirror Repeats (MR) → Triplex**: Form three-stranded DNA structures
- Unusual base pairing (Hoogsteen pairs)
- Regulate gene expression

**A-Phased Repeats (APR)**: Periodic A-tracts that bend DNA
- Affect nucleosome positioning
- Important for DNA bending and flexibility

**Short Tandem Repeats (STR)**: Microsatellites with repeated sequences
- Highly variable markers
- Used in forensics and population genetics

**Z-DNA**: Left-handed helix (vs. right-handed B-DNA)
- Rare but important in regulation
- Found at promoters and other regulatory regions

### Why This Matters

Non-B DNA structures are associated with:
- **Cancer**: Genomic instability, mutations
- **Neurological disorders**: Repeat expansions (Huntington's, Fragile X)
- **Aging**: Telomere shortening, G-quadruplex formation
- **Evolution**: Source of genetic variation

## References

For more information, see:

1. **Original GFA Publication:**
   - Cer RZ, Donohue DE, Mudunuri US, et al. (2013). Non-B DB v2.0: a database of predicted non-B DNA-forming motifs and its associated tools. *Nucleic Acids Research*, 41(D1):D94-D100. [doi:10.1093/nar/gks955](https://doi.org/10.1093/nar/gks955)

2. **Non-B DNA Database:**
   - https://nonb-abcc.ncifcrf.gov/

3. **Key Review Articles:**
   - Bacolla A, Tainer JA. (2019). Untangling the causes of non-B DNA structure-mediated genomic instability. *Trends in Biochemical Sciences*, 44(9):713-728.

## Session Information

```{r session-info}
sessionInfo()
```
